---
title: "Caste, Landownership, and Household Expenditure in Uttar Pradesh (ACCESS 2018)"
author: "Pranav Mehta"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    df_print: tibble
    fig_caption: true
---

# Introduction

Economic disparities across caste groups remain a central feature of India’s political economy. Prior research suggests that households belonging to the “General” caste category disproportionately own productive assets—particularly agricultural land—relative to their population share. This project uses the ACCESS 2018 survey to examine whether caste-based differences in monthly household expenditure persist in Uttar Pradesh, and whether landownership, gender, and educational attainment help explain observed gaps.

## Data and Sample Construction

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
```

```{r}
# ACCESS 2018 data must be placed locally in /data
load(here("data", "access2018.RData"))
```

```{r clean-data}
access <- access2018 %>%
  mutate(
    gender.fac    = factor(gender),
    caste.fac     = factor(caste),
    incsource.fac = factor(incsource),
    state.fac     = factor(state),
    education.fac = factor(education)
  ) %>%
  select(-HHID, -date)

access_up <- access %>%
  filter(state.fac == "UTTAR PRADESH") %>%
  mutate(
    reserved = factor(if_else(
      caste.fac %in% c("Other Backward Class",
                       "Scheduled Caste",
                       "Scheduled Tribe"),
      "Reserved", "General"
    ))
  )
```

Unit of observation: individual household heads or senior members.
Sample size (Uttar Pradesh): `r nrow(access_up)` individuals.

# Descriptive Statistics

```{r}
access_up %>%
  group_by(reserved) %>%
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    mean_monthly_exp = mean(month_exp, na.rm = TRUE),
    median_monthly_exp = median(month_exp, na.rm = TRUE),
    .groups = "drop"
  )
```

General caste households in Uttar Pradesh spend more on average than Reserved caste households, with a mean monthly expenditure gap of approximately Rs. 1,070.

# Landownership and Caste Representation

Households are defined as landowners if their primary income source is Agriculture (own land).

```{r landowners}
land_owners <- access_up %>%
  filter(incsource.fac == "Agriculture (own land)")
```

```{r}
pop_share <- prop.table(table(access_up$reserved))
land_share <- prop.table(table(land_owners$reserved))

tibble(
  caste_group = names(pop_share),
  population_share = as.numeric(pop_share),
  landowner_share = as.numeric(land_share),
  overrepresentation_ratio = landowner_share / population_share
)
```

General caste households constitute approximately 24% of the sample but account for 33% of landowning households, indicating substantial over-representation among landowners.

# Expenditure Gaps Among Landowners

```{r landowner-gap}
land_owners %>%
  group_by(reserved) %>%
  summarise(
    mean_monthly_exp = mean(month_exp, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```

Restricting the sample to landowning households reduces the General–Reserved expenditure gap, but does not eliminate it.

# Gender and Education Heterogeneity

## Gender

```{r}
land_owners %>%
  group_by(reserved, gender.fac) %>%
  summarise(
    mean_monthly_exp = mean(month_exp, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```

## Education (High School or Above)

```{r education-gap}
land_owners %>%
  filter(education.fac %in% c("Grade 10", "Grade 12", "Graduate or above")) %>%
  group_by(reserved) %>%
  summarise(
    mean_monthly_exp = mean(month_exp, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```

Higher educational attainment substantially narrows the caste-based expenditure gap, though differences remain.

# Discussion and Limitations

This analysis documents persistent caste-based expenditure gaps in Uttar Pradesh. Landownership and education explain a meaningful portion of the disparity, but do not fully account for it. Interpretation should be cautious: the ACCESS data lack survey weights and detailed sampling documentation, limiting external validity. Future work could incorporate multivariate regression analysis and state-level comparisons.

# Acknowledgements

This project was adapted and extended from coursework for U6614: Data Analysis for Policy Research using R, taught by Prof. Harold Stolper at Columbia SIPA. All analysis and code are my own.

